---
title: "Estimation and testing of vaccine sieve effects using machine learning"
author: "Blinded"
date: "October 6, 2016"
output:
    md_document:
        variant: markdown_github
---
    
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation of `survtmle` package
A beta verion of an `R` package is currently available on GitHub and can be installed as follows. 

```{r, message = FALSE}
# need devtools package for github install
library(devtools)

# install from github
# install_github("linktobeadded")

# load library
library(survtmle)
```

The main function in the package is the `survtmle` function, which computes the cumulative incidence for specified failure type(s). See `?survtmle` for more details, including examples using simulated data.

## Mock RV144 analysis
Below, we produce an analysis of the mock RV144 data that are included with the `survtmle` package. See `?rv144` for more description of the data set. 

```{r}
# load the data from the package
data(rv144)

# look at the first 6 rows of data
head(rv144)
```

Below we implement code that estimates the cumulative incidence for both matched and mismatched infections in both the vaccine and treatment arms. For computational convenience, we use a simplified Super Learner library relative to the one used in the manuscript. If one desires to use the same library as in the manuscript, those functions are available in the `superlearnerfunctions_rv144.R` file and can be implemented by sourcing the code from that file and uncommenting the `SL.ftime` and `SL.ctime` lines below. 

```{r, message = FALSE}
# will need package rpart and randomforest
library(rpart)
library(randomForest)

# set a seed because superlearner uses cross validation
set.seed(12345)

# to use simple super learner library, use these lines
SL.lib <- c("SL.glm","SL.mean","SL.rpart")
# to use super learner library from paper, use these lines
# SL.lib <- c("SL.glm","SL.mean","SL.rpart.caret1","SL.rf.caret1")

# fit the tmle at the last time point using survtmle
tmle.fit <- survtmle(
    ftime = rv144$ftime, 
    ftype = rv144$ftype,
    trt = rv144$vax,
    t0 = 6, # final timepoint, 3 years post enrollment
    adjustVars = rv144[,4:10], # baseline covariates
    glm.trt = "1", # empirical estimates of treatment probabilities
    SL.ftime = SL.lib, 
    SL.ctime = SL.lib, 
    method = "mean", # for iterative mean-based tmle
    verbose = FALSE, 
    returnModels = TRUE # return super learner fits for use with timepoints()
 )

# take a look at the output
tmle.fit
```

We can now use a call to the `timepoints` function in the `survtmle` package to obtain estimates at the first five scheduled visits. 

```{r}
# call timepoints to estimate incidence at each timepoint
tp.fit <- timepoints(
    object = tmle.fit, 
    times = 1:6, 
    returnModels = FALSE
    )

# check out the results for a single timepoint, e.g. t = 3
tp.fit$t3
```

We now define a function `getVE` that we use to help transform the results from the cumulative incidence scale to the vaccine efficacy and vaccine sieve effects scale. The function is not yet included in the `survtmle` package, but may be in the future. 

```{r}
#' getVE
#' 
#' A function that uses output from a survtmle object to compute
#' vaccine efficacy and vaccine sieve effect estimates
#' 
#' @param out A \code{survtmle} object with two types of failure
#' and two treatment types. 
#' @param tol Level of incidence below which not to compute VE. 
#' If incidence is very small in the placebo arm the asymmetric scale of VE can
#' cause very small values of VE. 
#' @return A list of data.frames whose first row is the vaccine efficacy against 
#' type 1 endpoints, second row is vaccine efficacy against type 2 endpoints
#' and third row is vaccine sieve effect. In each row the first entry corresponds
#' to the effect estimate, the second and third to the lower and upper confidence 
#' limits, and the fourth to the p-value for the Wald test of the 
#' relevant hypothesis test (VE = 0 or VSE = 1).

getVE <- function(out, tol = 1e-10){
    fnOut <- lapply(out, function(rslt){
        # check for very small values of cumulative incidence
        # that lead to very large/small values of VE
        if(all(rslt$est[1:2] > tol)){ 
            # ve against type 1 endpoints
            ve1 <- 1 - rslt$est[2]/rslt$est[1]
            # relative risk on the log scale
            log.rr1 <- log(rslt$est[2]/rslt$est[1])
            # gradient used for se computation on log-scale
            a <- matrix(c(-1/rslt$est[1], 1/rslt$est[2]),nrow=2)
            # standard error on the log scale for relative risk
            se.log.rr1 <- sqrt(t(a)%*%rslt$var[1:2,1:2]%*%a)
            # confidence interval
            ve1.low <- 1 - exp(log.rr1 + 1.96*se.log.rr1)
            ve1.high <- 1 - exp(log.rr1 - 1.96*se.log.rr1)
            # two sided p-value
            p1 <- 2*pnorm(-abs(log.rr1/se.log.rr1))
        }else{
            ve1 <- ve1.low <- ve1.high <- p1 <- NA
        }
        
        if(all(rslt$est[3:4] > tol)){
           # type 2 proceed exactly as above
            ve2 <- 1 - rslt$est[4]/rslt$est[3]
            log.rr2 <- log(rslt$est[4]/rslt$est[3])
            a <- matrix(c(-1/rslt$est[3], 1/rslt$est[4]),nrow=2)
            se.log.rr2 <- sqrt(t(a)%*%rslt$var[(3:4),(3:4)]%*%a)
            ve2.low <- 1 - exp(log.rr2 + 1.96*se.log.rr2)
            ve2.high <- 1 - exp(log.rr2 - 1.96*se.log.rr2)
            p2 <- 2*pnorm(-abs(log.rr2/se.log.rr2))  
        }else{
            ve2 <- ve2.low <- ve2.high <- p2 <- NA
        }
        
        if(all(rslt$est > tol)){
           # vaccine sieve effect
            sieve <- exp(log.rr2 - log.rr1)
            # on the log scale
            log.sieve <- log.rr2 - log.rr1
            # gradient used to compute standard error on the log scale
            a <- matrix(c(1/rslt$est[1], -1/rslt$est[2],
                          -1/rslt$est[3], 1/rslt$est[4]),nrow=4)
            # standard error
            se.log.sieve <- sqrt(t(a)%*%rslt$var%*%a)
            # confidence interval
            sieve.low <- exp(log.sieve -1.96*se.log.sieve)
            sieve.high <- exp(log.sieve +1.96*se.log.sieve)
            # p-value for hypothesis test
            sieve.p <- 2*pnorm(-abs(log.sieve/se.log.sieve))
        }else{
            sieve <- sieve.low <- sieve.high <- sieve.p <- NA
        }
        
        # return a matrix
        tmp <- data.frame(
            rbind(c(ve1, ve1.low, ve1.high, p1),
                  c(ve2, ve2.low, ve2.high, p2),
                  c(sieve, sieve.low, sieve.high, sieve.p)
        ))
        row.names(tmp) <- c("ve1","ve2","vse")
        colnames(tmp) <- c("est","ci.low","ci.high","p")
        return(tmp)
    })
    fnOut
}

# get the ve and vse 
ve.fit <- getVE(tp.fit)

# look at the results at time 6
ve.fit$t6
```

## Plotting the results
Below is the plot code used to generate the manuscript figures.

```{r, cache = TRUE}
# figure dimensions -- used to compute appropriate margins
figure.width <- 8
figure.height <- 5.5
# computing appropriate margins
par(mar=c(4.1,0,0,0))
left.1 <- par()$mai[1]
par(mar=c(4.1,0,0,0))
right.1 <- par()$mai[1]
left.2 <- right.2 <- right.1
bottom.1 <- upper.2 <- right.1
bottom.2 <- upper.1 <- left.1
thisCex <- 1
x <- (figure.width - (left.1 + right.1 + left.2 + right.2))/2
wid <- rep(c(1,(x+left.2+right.2)/(x+left.1+right.1)),2)
x <- (figure.height - (upper.1 + upper.2 + bottom.1 + bottom.2))/2
hei <- c((x+upper.2+bottom.2)/(x+upper.1+bottom.1),1,1,(x+upper.2+bottom.2)/(x+upper.1+bottom.1))

# set up the layout
layout(matrix(c(1,3,2,4),2,2),width=wid)
# make the top row of plots
for(j in 1:2){
  if(j==2){
    par(mar=c(2.1,4.1,4.1,2.1))
    main.tit <- ""
    yl <- ""
  }else if(j==1){
    par(mar=c(2.1,4.1,4.1,2.1))
    main.tit <- ""
    yl <- "Cumulative incidence"
  }
  # get vector of cumulative incidence in vaccine arm
  f0.hat <- unlist(lapply(tp.fit, function(f){
      f$est[(2*j - 1)]
  }))
  # get vector of cumualtive incidence in vaccine arm
  f1.hat <- unlist(lapply(tp.fit, function(f){
      f$est[(2*j)]
  }))
  
  plot(c(0,f0.hat)~I(0:6),cex.lab=1.2,pch="",col=1,
       ylim=c(0,0.010),bty="n",cex=thisCex,
       #     ylim=c(0,0.004),
       ylab=yl,
       xaxt="n",xlab="Years since entry",
       main=main.tit)
  segments(x0=0,x1=7, y0=seq(0,0.010,0.002),col="gray90")
  points(c(0,f0.hat)~I(0:6),pch=ifelse(j==2,19,17),col=1,cex=thisCex)
  points(c(0,f1.hat)~I(0:6),pch=ifelse(j==2,1,2),col=1,cex=thisCex)
  axis(side=1, at=seq(0,6,2),label=seq(0,3,1))
  legend(x="topleft",pch=ifelse(rep(j,2)==c(2,2),c(19,1),c(17,2)),col=1,
         bty="n",
         legend=c("Placebo","Vaccine"))
  mtext(side=3,paste0("AA site 169",ifelse(j==1," matched"," mismatched")))
}

#--------------------------
# vaccine efficacy plot
#--------------------------
# get matrix of vaccine efficacies, first column = matched
veMat <- Reduce("rbind", lapply(ve.fit,function(f){
    f$est[1:2]
}))

par(mar=c(4.1,4.1,2.1,2.1))
yl <- "Vaccine Efficacy"
y.lim <- c(-0.75,1.25)
plot(c(0,veMat[,1])~I(0:6),ylab=yl,col=1,
     xlab="Years since entry",cex.lab=1.2,xaxt="n",bty="n",cex=thisCex,
     ylim=y.lim,pch="")
segments(x0=0,x1=7, y0=seq(-1,1,0.5),col="gray90")
points(veMat[1:6,1]~I(1:6),ylab=yl,col=1,pch=17)
points(veMat[2:6,2]~I(2:6),ylab=yl,col=1,pch=19,lty=2)
axis(side=1, at=seq(0,6,2),label=seq(0,3,1))
legend(x=0,y=1.55,xpd=TRUE, pch=c(17,19),col=1,bty="n",
       legend=c("169 matched","169 mismatched"))

#-------------------------
# sieve effect plot
#-------------------------
# get vector of vse
vse <- Reduce("rbind", lapply(ve.fit,function(f){
    f["vse",1:3]
}))

par(mar=c(4.1,4.1,2.1,2.1))
yl <- "Sieve Effect"
y.lim <- c(0,12)

plot(c(0,0,vse[2:6,1])~I(0:6),ylab=yl,col=1,
     xlab="Years since entry",cex.lab=1.2,xaxt="n",bty="n",cex=thisCex,
     ylim=y.lim,pch="")
segments(x0=0,x1=7, y0=seq(0,10,2),col="gray90")
points(vse[2:6,1]~I(2:6),ylab=yl,col=1,pch=19)
axis(side=1, at=seq(0,6,2),label=seq(0,3,1))
abline(h=1,lty=3)
segments(x0=4:6,y0=vse[4:6,2],y1=vse[4:6,3])
arrows(x0=2:3,y0=vse[2:3,2],y1=12,length=0.075)
```

## Mock RTS,S/AS01 analysis
We now produce an analysis of the multiply outputed mock RTS,S/AS01 data that are included with the `survtmle` package. See `?rtss` for further description of the data set. 
```{r}
# load the data from the package
data(rtss)

# look at the first 6 rows of data for first data set
head(rtss[[1]])
```

For simplicity, we will focus on implementing the analysis at a single timepoint to illustrate how the multiple outputation procedure works. The `timepoints` function could easily be used to duplicate the analysis across every one month period. As with the RV144 analysis above, we will show code
illustrating the method using a simplified Super Learner library. If one desires to use the Super Learner library from the paper, those functions are available in the `superlearnerfunctions_rtss.RData` file. 

```{r, message = FALSE}
# needed packages 
library(rpart)
library(randomForest)
library(gbm)

# set a seed because superlearner uses cross validation
set.seed(12345)

# to use simple super learner library, use these lines
SL.lib <- c("SL.glm","SL.mean")
# to use super learner library from paper, use these lines
# SL.lib <- list(c("SL.glm","All"),
#                c("SL.gam","screen.corRank5"),
#                c("SL.gam","screen.corRank10"),
#                c("SL.gbm.itMeans","All"),
#                c("SL.randomForest.itMeans","All"),
#                c("SL.mean","All"),
#                c("SL.step.interaction","screen.corRank5"),
#                c("SL.step.forward","All"),
#                c("SL.glm","trtOnlyScreen"),
#                c("SL.glm","sitezScreen"),
#                c("SL.glm.interaction","sitezScreen"))


# initialize empty list for results
rtss.rslt <- vector(mode = "list", length = 10)
# fit the tmle at time 4 for each of the MO data sets
for(m in 1:10){
    # print messages to indicate progress
    if(m==1){
        cat("Estimating time 1")
    }else{
        cat(paste0(" ,",m))
    }
    
    rtss.rslt[[m]] <- survtmle(
        ftime = rtss[[m]]$ftime, 
        ftype = rtss[[m]]$ftype,
        trt = rtss[[m]]$vaccine,
        t0 = 4, # four months post-vaccination
        adjustVars = rtss[[m]][,4:ncol(rtss[[m]])], # baseline covariates
        glm.trt = "1", # empirical estimates of treatment probabilities
        SL.ftime = SL.lib, 
        SL.ctime = SL.lib, 
        method = "mean", 
        verbose = FALSE, 
        returnModels = FALSE # to save space
    )
}

# take a look at the output for first outputed data set
rtss.rslt[[1]]
```


The results can be combined through multiple outputation using the function `getMO`, which is not included in the `survtmle` package, but may be included at a future date.

```{r}
#' getMO
#' 
#' A function that performs multiple outputation point estimation and
#' confidence interval calculation. 
#' 
#' @param rsltList A \code{list} of output from \code{survtmle} at 
#' a single timpoint, where each entry in the list corresponds with 
#' the results for one outputed data set. 
#' @param N The number of rows in the data used to compute the 
#' \code{survtmle} output. This is needed for variance calculations.
#' @return A list of \code{data.frames} containing results for the
#' multiply outputed estimates of cumulative incidence, vaccine efficacy,
#' and vaccine sieve effects and corresponding confidence intervals. 
#' The sieve effect output also includes a p-value for the test of equal
#' vaccine efficacy between groups. 

getMO <- function(rsltList,
                  n = 6890){
    M <- length(rsltList)
    # initialize empty result lists
    FVector_Z0_J1 <- FVector_Z1_J1 <- rep(NA,M)
    FVector_Z0_J2 <-FVector_Z1_J2 <- rep(NA,M)
    varLogCumIncRatio_J2 <- varLogCumIncRatio_J1 <- rep(NA,M)
    varVector_Z0_J1 <- varVector_Z1_J1 <- rep(NA,M)
    varVector_Z0_J2 <- varVector_Z1_J2 <- rep(NA,M)
    Dmat_Z0_J1 <- Dmat_Z1_J1 <- matrix(NA,M,n)
    Dmat_Z0_J2 <- Dmat_Z1_J2 <- matrix(NA,M,n)
    
    ct <- 0
    for(i in 1:M){
        fit <- rsltList[[i]]
        ct <- ct+1
        # vector of estimates
        FVector_Z0_J1[ct] <- fit$est[1]
        FVector_Z1_J1[ct] <- fit$est[2]
        FVector_Z0_J2[ct] <- fit$est[3]
        FVector_Z1_J2[ct] <- fit$est[4]
        # matrix of influence curves
        Dmat_Z0_J1[ct,] <- fit$ic[,1]
        Dmat_Z1_J1[ct,] <- fit$ic[,2]
        Dmat_Z0_J2[ct,] <- fit$ic[,3]
        Dmat_Z1_J2[ct,] <- fit$ic[,4]
    }

    #------------------------------
    # estimates of cumulative inc. 
    #------------------------------
    # point estimates
    cumInc_Z0_J1_MO <- mean(FVector_Z0_J1)
    cumInc_Z1_J1_MO <- mean(FVector_Z1_J1)
    cumInc_Z0_J2_MO <- mean(FVector_Z0_J2)
    cumInc_Z1_J2_MO <- mean(FVector_Z1_J2)

    # variance estimates
    a <- rep(1,M)
    varIC_Z0_J1_MO <- varIC_Z1_J1_MO <- 0
    varIC_Z0_J2_MO <-  varIC_Z1_J2_MO <- 0
    for(i in 1:n){
        varIC_Z0_J1_MO <- varIC_Z0_J1_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%Dmat_Z0_J1[,i])
        varIC_Z1_J1_MO <- varIC_Z1_J1_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%Dmat_Z1_J1[,i])
        varIC_Z0_J2_MO <- varIC_Z0_J2_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%Dmat_Z0_J2[,i])
        varIC_Z1_J2_MO <- varIC_Z1_J2_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%Dmat_Z1_J2[,i])
    }

    #---------------------------------------
    # estimates for log cum. inc. ratio
    #---------------------------------------
    # point estimates
    logCumIncRatio_J1_MO <- mean(log(FVector_Z1_J1/FVector_Z0_J1))
    logCumIncRatio_J2_MO <- mean(log(FVector_Z1_J2/FVector_Z0_J2))

    # variance estimates
    a <- rep(1,M)
    gMatrix_J1 <- t(cbind(-1/FVector_Z0_J1, 1/FVector_Z1_J1))
    gMatrix_J2 <- t(cbind(-1/FVector_Z0_J2, 1/FVector_Z1_J2))

    varLogCumIncRatio_J1_MO <- varLogCumIncRatio_J2_MO <- 0
    for(i in 1:n){
        thisD_J1 <- rbind(Dmat_Z0_J1[,i],Dmat_Z1_J1[,i])
        thisD_J2 <- rbind(Dmat_Z0_J2[,i],Dmat_Z1_J2[,i])
        thisMat1 <- colSums(gMatrix_J1*thisD_J1)
        thisMat2 <- colSums(gMatrix_J2*thisD_J2)
        varLogCumIncRatio_J1_MO <- varLogCumIncRatio_J1_MO + 
           1/n^2 * tcrossprod(1/M * t(a)%*%thisMat1)
        varLogCumIncRatio_J2_MO <- varLogCumIncRatio_J2_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%thisMat2)
    }
    
    #------------------------------
    # estimates for sieve effects
    #------------------------------
    # point estimates
    logRatioCumIncRatio_MO <- mean(
        log(FVector_Z1_J1/FVector_Z0_J1 / (FVector_Z1_J2/FVector_Z0_J2))
    )
    # variance estimates
    a <- rep(1,M)
    gMatrix <- t(cbind(
        -1/FVector_Z0_J1,1/FVector_Z1_J1, 1/FVector_Z0_J2,-1/FVector_Z1_J2
      ))

    varLogRatioCumIncRatio_MO <- 0
    for(i in 1:n){
        thisD <- rbind(Dmat_Z0_J1[,i],Dmat_Z1_J1[,i],
                       Dmat_Z0_J2[,i],Dmat_Z1_J2[,i])
        thisMat <- colSums(gMatrix*thisD)
        varLogRatioCumIncRatio_MO <- varLogRatioCumIncRatio_MO + 
            1/n^2 * tcrossprod(1/M * t(a)%*%thisMat)
    }
    
    #------------------------------
    # format results
    #------------------------------
    out <- vector(mode="list",length=0L)
  
    out$cumInc <- data.frame(Z=c(0,1,0,1),J=c(1,1,2,2),
                               cumInc=c(cumInc_Z0_J1_MO, 
                               cumInc_Z1_J1_MO, 
                               cumInc_Z0_J2_MO, 
                               cumInc_Z1_J2_MO))

    out$ve <- data.frame(
          J=c(1,2),
          ve=c(1-exp(logCumIncRatio_J1_MO),
               1-exp(logCumIncRatio_J2_MO)),
          ve.low=c(1 - exp(logCumIncRatio_J1_MO +
                               1.96*sqrt(varLogCumIncRatio_J1_MO)),
                   1 - exp(logCumIncRatio_J2_MO +
                               1.96*sqrt(varLogCumIncRatio_J2_MO))),
          ve.high=c(1 - exp(logCumIncRatio_J1_MO -
                                1.96*sqrt(varLogCumIncRatio_J1_MO)),
                    1 - exp(logCumIncRatio_J2_MO -
                                1.96*sqrt(varLogCumIncRatio_J2_MO)))
          )
    
    out$sieve <- data.frame(
        ratio=exp(logRatioCumIncRatio_MO),
        ratio.low=exp(logRatioCumIncRatio_MO -
                         1.96*sqrt(varLogRatioCumIncRatio_MO)),
        ratio.high=exp(logRatioCumIncRatio_MO +
                           1.96*sqrt(varLogRatioCumIncRatio_MO)),
        p=2*pnorm(-abs(logRatioCumIncRatio_MO/
                           sqrt(varLogRatioCumIncRatio_MO)))
        )
  
  return(out)
}
```

We can use this function to obtain results on the simulated RTS,S/AS01 data. 
```{r}
# perform MO over the 10 simulated data sets
mo.rslt <- getMO(rsltList = rtss.rslt, n = 6890)
# look at the results
mo.rslt
```




